---
title: "Guide to Water Quality Figure"
author: "Cory Sauve"
date: "10/7/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
wq7 <- readr::read_csv(here::here("5_output", "wq7.csv"))
secchi <- 0.75
one_per_light <- 2.1
```


# Purpose 

The following guide is to serve as a walk through to build the *framework* for Figures 1-5. It will focus on Figure 1 (Temp, DO, % Sat, Turbidity).  Once you get a handle on how to make the first figure, you can easily apply the concepts to complete Figures 2-5.  You can view all of the required figures in the *Required figures for reports* on Canvas.  

# Required Libraries 

We're going to use the `ggplot2` package to make all of our figures.  It's already included in the `tidyverse` package we have already been using.  We will also use `cowplot` to turn individual figures into gridded figures. 
```{r}
library(tidyverse)
library(cowplot)
```

# Required Data 

You should already have completed the *Limno Data Manipulation Guide* and have the dataframe `wq7` loaded into your environment.  If you do not have `wq7` ready, go back and complete the manipulation guide.  `wq7` should look the this:
```{r, echo=TRUE, eval=FALSE}
View(wq7)
```

# Temperature Figure 

The first figure we'll make is the temperature figure. I'll spend a little more time explaining each step to make this figure so you can get an idea of how `ggplot2` works.

## Creating a new `ggplot`

The first step to creating a figure with `ggplot2` is to create a `ggplot`.  We'll use the function `ggplot()` to build the initial framework of our figure.  `ggplot()` requires two arguments: the data and the aesthetics.  You can this of aesthetics as the variables that represent the x and y-axis variables.  Here we will use the `wq7` dataframe and then "map" `temp_c` as the x-variable and `depth` as the y-variable.  

```{r, eval=TRUE, echo=TRUE}
ggplot(data = wq7, aes(x = temp_c, y = depth))
```

Notice that we have now created the base framework for our plot, where we see that temperature is now on the x-axis and depth is on the y-axis.  However, we will not see any data until we add additional *layers* to our figure, known as `geoms`. 

## Adding data layers with `geom`

Now that we have the foundation of our temperature figure using `ggplot`, we'll add two layers to add points and lines to the figure.  `ggplot2` uses a `+` sign to add layers to a figure.  It is easy to mix this up with the `%>%` operator (I do it ALL the time).  So if you get a weird error message I'd look there first! We can add points with `geom_point()` and lines with `geom_path()`. The `size = ` argument inside the functions will control the point and line size. 
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1)
```

## Adding error bars 

We now have data plotted! To add error bars for the duplicate and replicate samples we'll pass the `geom_errorbarh()` function.  Note the addition of `h` at the end of the function.  If you omit that and just use `geom_errorbar()` you'll end up with vertical error bars.  `geom_errorbarh` requires several arguments inside of the `aes()` function.  We need to define the y-variable and then set the bounds for the length of the error bars on the x-axis.  We will do this as follows:
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) 
```

## Annotating the Secchi depth

We want to include a line indicating the depth of our Secchi measurement, and then include a text annotation.  We can do this with `geom_segment()` and `geom_text()`. This will take some trial and error to place the line and text where we want them.  We want to center the line on the a-axis so having the line from 15-20 seems like a logical place to start.  We also need to define where the line will be on the y-axis, and will use the previously saved object `secchi` to do so.  Finally, we'll define the type of line we want.  In this case, we want a dashed line. 
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") 
```

`geom_text()` will now allow us to label the line
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) 
```

## Changing the appearance 

We are getting close to having a completed figure! The final steps are to add scale arguments and change the overall *theme* of the figure.  `coord_cartesian()` allows us to manually set the extent of the x and y-axes. You will want to look at the range of the data to determine an appropriate scale.  Here we will set the x-axis range from 9-26 and leave the y-axis be.  
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) +
  coord_cartesian(xlim = c(9, 26)) 
```

Now we will change the axis labels.  You need to know the type of data you have in order to do this correctly.  We are dealing with continuous data so we will pass `scale_x_continuous()` to set the x-axis.  However, we want to flip the y-axis so we will pass `scale_y_reverse()` to do so.
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) +
  coord_cartesian(xlim = c(9, 26)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Temperature (°C)") 
```

We now have a figure that is about 95% completed.  The final step is to change the overall appearance of the figure with a `theme()`.  `ggplot2` comes with a bunch of pre-made themes.  However, we will use a custom theme called `theme_awesome()`.  It is not important that you know what every line in the function does besides that it will modify the figure's appearance 
```{r}
# Define ggplot2 theme for plots 
theme_awesome <- function(){
  theme(
    plot.background = element_rect(fill = "gray100", color = NA),
    panel.background = element_rect(fill = "gray96", color = NA),
    panel.border = element_rect(fill = NA, color = "grey90"),
    panel.grid.major = element_line(color = "grey90", size = 0.5),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "grey25", size = 1),
    axis.line.y = element_blank(),
    axis.text = element_text(color = "grey60", size = 10, family = "Helvetica"),
    axis.title = element_text(color = "grey25", size = 12),
    axis.ticks.x = element_line(size = 0.5, color = "grey25"),
    axis.ticks.y = element_line(size = 0.5, color = "grey90"),
    axis.ticks.length = unit(0.15, "cm")
  )
}
```

Simply add `theme_awesome()` to the end of the code used to create the above figure. 
```{r}
ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) +
  coord_cartesian(xlim = c(9, 26)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Temperature (°C)") +
  theme_awesome()
```

Congrats! You just finished 1/4 of the first figure.  I know that seemed like a lot of code, but we can essentially copy and past the code to create the remaining sub-plots in Figure 1.  We just need to change the variables and labels. We will want to save the above plot as an object as follows:
```{r}
p_temp <- ggplot(data = wq7, aes(x = temp_c, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_sd, xmax = temp_c + temp_sd), height = 0.2) +
  geom_segment(aes(x = 15, y = secchi, xend = 20, yend = secchi), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) +
  coord_cartesian(xlim = c(9, 26)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Temperature (°C)") +
  theme_awesome()
```

# The remaining figures 

I am not going to go over how to create the remaining sub-plots in much detail.  Just follow the above steps and change variables when necessary.  

## D.O. figure 
```{r do, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
p_do <- ggplot(data = wq7, aes(x = do_mgl, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = do_mgl - do_sd, xmax = do_mgl + do_sd), height = 0.2) +
  geom_text(aes(x = 8, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 6, y = one_per_light, xend = 10, yend = one_per_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 16)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "Dissolved Oxygen (mg/L)", breaks = seq(0, 16, 4)) + 
  theme_awesome()
p_do
```

## % SAT figure 
```{r sat, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
p_do_sat <- ggplot(data = wq7, aes(x = do_sat_per, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = do_sat_per - do_sat_sd, xmax = do_sat_per + do_sat_sd), height = 0.2) +
  coord_cartesian(xlim = c(0, 200)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "D.O. Saturation (%)", breaks = seq(0, 200, 40)) +
  theme_awesome()
p_do_sat
```

## Turbidity figure 
```{r turb, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
p_turb <- ggplot(data = wq7, aes(x = turb_ntu, y = depth)) +
  geom_point(size = 2) +
  geom_path(size = 1) +
  geom_errorbarh(aes(y = depth, xmin = turb_ntu - turb_sd, xmax = turb_ntu + turb_sd), height = 0.2) +
  coord_cartesian(xlim = c(10, 40)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Turbidity (NTU)", breaks = seq(10, 40, 5)) +
  theme_awesome()
p_turb
```

## Putting them together to create Figure 1

Now that we have the four sub-plots completed and stored as individual object, we can put them in a grid to complete Figure 1.  We will use the `plot_grid()` function in the `cowplot` package to do this. `plot_grid()` requires several arguments: the plot object we saved above, how many rows we want, the overall scale, text labels, and then the size of the labels.  
```{r plot1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
plot1 <- plot_grid(p_temp, p_do, p_turb, p_do_sat, nrow = 2, scale = 1, labels = c("A.", "B.", "C.", "D."), label_size = 12)
plot1
```

We can see that we now have the four sub-plots in a 2x2 grid. Make sure that you save the new plot as an object with the `<-` operator! 

## Saving the figure 

The final step is to save the figure.  We will use the `ggsave()` function to do this. You will need to first define what plot to save, name the file, and then control the dimensions with `height =` and `width =`.  
```{r}
ggsave(plot1, file = "plot1.png", height = 7, width = 7)
```

If the function worked, you should see a new plot in your directory! 