---
title: "Coding Basics"
author: "Cory Sauve"
date: "9/7/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Libraries used 
```{r libraries, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## The Basics of R 

#### Coding Bacics 

You can use R as a (absurdly powerful) calculator...
```{r calculator, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
1 + 9 
1 / 200 * 30
(59 + 73 + 2) / 3 
3 ^ 4
3 ** 4 # this way is not a good idea, but works :/
```

R also supports all the basic log and trig functions you can find on a scientific calculator...
```{r log and trig, echo = TRUE, warning = FALSE, message = FALSE}
sqrt(2)
sin(pi / 2)
log(4) # natural log
log10(4) # common log 
```

If you want to see what a function does, simply call the `?` in front of a function.
```{r function help, echo = TRUE, warning = FALSE, message = FALSE}
?log()
```

The ability to create *objects* is an essential component of R.  You can create an object with `<-`:
```{r objects numeric, echo = TRUE, warning = FALSE, message = FALSE}
x <- 3 * 4
x
```

We can also assign a character string to an object
```{r objects char, echo = TRUE, warning = FALSE, message = FALSE}
my_string <- "R is kinda cool" 
```

We can do math with objects and numbers 
```{r object math, echo = TRUE, warning = FALSE, message = FALSE}
x * 2
```

Or we can do math with multiple objects 
```{r object math 2, echo = TRUE, warning = FALSE, message = FALSE}
one_fish <- 1
two_fish <- 2
one_fish + two_fish
```

We can also use math to make other objects 
```{r object math 3, echo = TRUE, warning = FALSE, message = FALSE}
three_fish <- one_fish + two_fish
three_fish
```

However, we can only do math with objects that are numeric 
```{r object fail, eval = FALSE, echo = TRUE, warning = FALSE, message = FALSE}
my_string + 2
```

#### What's in a Name? 

Object names must start with a letter, and can only contain letters, numbers, _, and **.**.  It is important to make descriptive object names and maintain a similar convention.  Some common (and uncommon) conventions include:  

  `some_people_like_snake_case`
  `othersUseCamelCase`
  `some.like.periods`
  `And_aFew.Others_live_OnTheEdge`

#### R is picky...
```{r picky, echo = TRUE, warning = FALSE, message = FALSE}
CodingIsRad <- 2 ^ 3
```

```{r picky2, eval = FALSE, echo = TRUE, warning = FALSE, message = FALSE}
CodingisRad
```

#### Coding Challenge 

Let's say you are working as a intern with IDEM and you conducted an electrofishing survey on Clear Creek.  They want you to calculate the catch-per-unit-effort (CPUE) for the fishes you caught. Catch-per-unit-effort for a given species can be calculated simply as $CPUE = Individuals caught / time$. Calculate CPUE for each species collected in hours. The data you collected are as follows:

```{r basics problem, eval = TRUE, echo = FALSE, warning = FALSE, message = FALSE}
catch_data <- data.frame(spp = c("Largemouth bass", "Bluegill", "Central stoneroller"),
                   num_collected = c(50, 15, 60),
                   effort_min = c(10, 10, 10)
              )
```

```{r basics problem 2, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
catch_data
```

Solution #1 - The hard way 
```{r basics problem sol 1, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
MISA_num <- 50
MISA_effort_hr <- 10 / 60
MISA_cpue <- MISA_num / MISA_effort_hr
MISA_cpue
LEMA_num <- 15 
LEMA_effort_hr <- 10 / 60
LEMA_cpue <- LEMA_num / LEMA_effort_hr
LEMA_cpue
CAAM_num <- 60
CAAM_effort_hr <- 10 / 60
CAAM_cpue <- CAAM_num / CAAM_effort_hr
CAAM_cpue
```

Solution #2 - The elegant way 
```{r basics problem sol 2, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
cpue <- function(num_col, effort_min){
  # Calculates catch-per-unit-effort for a species given catch and effort (minutes)
  effort_hr <- effort_min / 60
  cpue <- num_col / effort_hr
  return(cpue)
}
catch_data %>% 
  mutate(
    cpue_hr = cpue(num_collected, effort_min)
  )
```

## Importing data

R can handle a wide variety of file types. I personally recommend using .csv format, or comma separated value, when importing data into R.  It's by far the most common in data analysis relating to environmental science.  That being said, R can also handle a wide variety of file types: Excel files (.xls, .xl), tab-separated files (.tsv), general delimited files (.delim) to name a few.  

Many of you probably use Excel to enter data (and probably analyze it too). You can easily export Excel files into .csv files when you go to save a file.  This is probably the best approach, and the one in which we will work with.  If you want to use Excel files, I suggest using the `readxl` package to do so.  For us, we will use the `readr` package.  The `readr` package is fast to use and already comes in when you call the `tidyverse` package (e.g. `library(tidyerse)`). We will use the `read_csv` function to load our .csv file into R. 
```{r read_csv, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
?read_csv
```

#### The Mighty %>%

Using a pipe (`%>%`) in R is an incredibly powerful tool. The `%>%` allows you to write code that reads "left to right" instead of from "inside out".  Technically, the `%>%` is found in the `magrittr` package but is ready for use from the `tidyverse`

To show how pipes work, let's look at a comparison on how code looks without and with a pipe.  Let's first create a vector of random values and call it x.  Note the use of the `c()` (combine) function here. The use of `c()` is essential to being able to program in R. 
```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
x <- c(0.109, 0.359, 0.63, 0.996, 0.515, 0.142, 0.017, 0.829, 0.907)
```

Now let's say that we want to exponentiation each value in `x`, sum those values, and then round them to a whole number.  The pipe-less way of doing so would look like this:
```{r,eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
round(sum(exp(x)))
```

Using a pipe allows us to take these functions and start a logical progression from left (start) to right (finish)
```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
x %>% 
  exp() %>% 
  sum() %>% 
  round()
```

#### `%>%` Coding Challenge 

Take a few minutes and complete the following tasks.  Create two solutions: one without using a pipe, and the other with pipes.  

1) Create a vector of numbers from 1-10 
2) Take the vectors of numbers, calculate the log for each number, then sum the numbers, and finally round to one decimal place (hint: use `?` to inspect `round()` to see if it includes something to handle the decimal places). Do this in one line of code without using pipes. 
3) Do the same thing using pipes 

Without pipes 
```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
x <- c(1:10)
round(sum(log(x)), digits = 1)
```

With pipes 
```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
x %>% 
  log() %>% 
  sum() %>% 
  round(digits = 1)
```

##### Combining values with `c()`

Combining values into a vector is an essential function in R.  Vectors are the building block of dataframes and R is very good at doing vector math.  As you get more advanced in R, vectors will constitute the majority of you analysis.  

To create a vector, we can use the `c()` function.  We won't go over every datatype that a vector can support, but you'll mostly come across vectors with numeric or character data. Let's make a vector of numeric values between 1-10, called `x`. 

```{r}
a_numeric_vector <- c(1:10)
a_numeric_vector
```

A vector of characters 
```{r}
a_char_vector <- c("this", "is", "a", "character", "vector")
a_char_vector 
```