---
title: "Guide to Plankton Figure"
author: "Cory Sauve"
date: "10/4/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
aphanizomenon <- readr::read_csv(here::here("5_output", "aphanizomenon.csv"))
bosmina <- readr::read_csv(here::here("5_output", "bosmina.csv"))
calanoid <- readr::read_csv(here::here("5_output", "calanoid.csv"))
ceratium <- readr::read_csv(here::here("5_output", "ceratium.csv"))
chaborus <- readr::read_csv(here::here("5_output", "chaborus.csv"))
cyclopoid <- readr::read_csv(here::here("5_output", "cyclopoid.csv"))
dolichospermum <- readr::read_csv(here::here("5_output", "dolichospermum.csv"))
microcystis <- readr::read_csv(here::here("5_output", "microcystis.csv"))
nauplii <- readr::read_csv(here::here("5_output", "nauplii.csv"))
one_per_light <- 2.1
secchi <- 0.75
```

# Purpose 

The purpose of this guide is to go over how to make the phytoplankton and zooplankton figures.  This is a fair warning that these two plots require the most code and are the most complex.  Make sure you do not wait until the last minute to put these together! 

# Required libraries 

We're going to use the `ggplot2` package to make all of our figures.  It's already included in the `tidyverse` package we have already been using.  
```{r}
library(tidyverse)
library(data.table)
```

# Required data 

You should already have completed the *Limno Data Manipulation Guide* and have the `plankton` dataframe in you environment along with the 9 dataframes for each genus we are going to plot.  If you do not have these, you will need to go through the data manipulation guide.  

# Creating additional dataframes 

We need to create some additional dataframes in order to create the two plankton figures.  These will allow us to add various elements to our figures.  It is not important that you know how the following two code chunks work.  However, if you get an error message from running the code you probably don't have the `data.table` package installed correctly or loaded in your environment.  Use `install.packages()` to install `data.table`, and then load it with `library()`.  You can also call the package with `data.table::` to avoid using the `library()` function.  Note that I chose a random number to indicate the bottom of the epilimnion in `epi_bot` and `epi_bot2`.  You need to measure this yourself using the materials in *Wetzel* or with Prof. Royer's help! 

**Phytoplankton**
```{r plankton df, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
one_per <- data.table::data.table(taxa = c("Aphanizomenon", "Ceratium", "Dolichospermum", "Microcystis"),
                 value = c(one_per_light, one_per_light, one_per_light))
one_per_name <- data.table::data.table(taxa = c("Aphanizomenon", "Ceratium", "Dolichospermum", "Microcystis"),
                                       name = c("", "", "", "1% Light Level (2.1m)"))
epi_bot <- data.table::data.table(taxa = c("Aphanizomenon", "Ceratium", "Dolichospermum", "Microcystis"),
                                  value = c(1.5, 1.5, 1.5, 1.5))
epi_bot_name <- data.table::data.table(taxa = c("Aphanizomenon", "Ceratium", "Dolichospermum", "Microcystis"),
                                       name = c("", "", "", "Bottom of Epi. (1.5m)"))
cha_presence <- data.table::data.table(taxa = c("Aphanizomenon", "Ceratium", "Dolichospermum", "Microcystis"),
                                       name = c("", "", "", "X"))
phyto_plotting <- rbind(dolichospermum, aphanizomenon, microcystis, ceratium)
```

**Zooplankton**
```{r, message=FALSE, warning=FALSE}
one_per2 <- data.table::data.table(taxa = c("Bosmina", "Calanoid", "Cyclopoid", "Nauplii"),
                 value = c(one_per_light, one_per_light, one_per_light))
one_per_name2 <- data.table::data.table(taxa = c("Bosmina", "Calanoid", "Cyclopoid", "Nauplii"),
                                       name = c("", "", "", "1% Light Level (2.1m)"))
epi_bot2 <- data.table::data.table(taxa = c("Bosmina", "Calanoid", "Cyclopoid", "Nauplii"),
                                  value = c(1.5, 1.5, 1.5, 1.5))
epi_bot_name2 <- data.table::data.table(taxa = c("Bosmina", "Calanoid", "Cyclopoid", "Nauplii"),
                                       name = c("", "", "", "Bottom of Epi. (1.5m)"))
cha_presence2 <- data.table::data.table(taxa = c("Bosmina", "Calanoid", "Cyclopoid", "Nauplii"),
                                       name = c("", "", "", "X"))
zoo_plotting <- rbind(bosmina, nauplii, calanoid, cyclopoid)
```

# Another custom ggplot2 theme 

We need to create a new function for the custom theme of the plankton figures.  It is very similar to the `theme_awesome()` theme we made for other figures.  I know some of you had issues with this function working correctly.  If it doesn't work, please send me an email or come to my office hours to troubleshoot.  
```{r theme 2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
theme_plankton <- function(){
  theme(
    plot.background = element_rect(fill = "gray100", color = NA),
    panel.background = element_rect(fill = "gray96", color = NA),
    panel.border = element_rect(fill = NA, color = "grey90"),
    panel.grid.major = element_line(color = "grey90", size = 0.5),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "grey25", size = 1),
    axis.line.y = element_blank(),
    axis.text = element_text(color = "grey60", size = 10, family = "Helvetica"),
    axis.title = element_text(color = "grey25", size = 12),
    axis.ticks.x = element_line(size = 0.5, color = "grey25"),
    axis.ticks.y = element_line(size = 0.5, color = "grey90"),
    axis.ticks.length = unit(0.15, "cm"),
    strip.background = element_rect(color = "white", fill = "white", size = 1, linetype = "solid"),
    strip.text.x = element_text(size = 10, color = "grey25", face = "italic")
  )
}
```

# Phytoplankton Plot 

We will use many of the same concepts we used to make the other figures.  However, we have to go through some additional steps to make all of the figures on the same plot.  

## The `ggplot` framework 

We will first pass an *empty* `ggplot()` argument and then add our variables to `geom_path()` and `geom_point` individually. We will also add error bars the same way we have done previously with `geom_errorbarh()`.
```{r}
ggplot() +
  geom_point(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 2) +
  geom_path(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 1) +
  geom_errorbarh(data = phyto_plotting, aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.2) 
```

## Adding error bars 

The output from the previous code will look terrible! Don't worry...this is what we wanted.  Essentially, we plotted all four variables on the same axis. To make them separate plots, we can use `facet_wrap()` and split by the variable `taxa`. Made sure you don't forget the `~` in front of the variable and specify one row with `nrow = 1`! 
```{r}
ggplot() +
  geom_point(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 2) +
  geom_path(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 1) +
  geom_errorbarh(data = phyto_plotting, aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.2) +
  facet_wrap(~taxa, nrow = 1) 
```

## Modifying the axes 

Our next step is to modify the x and y-axis.  We need to set our x-axis limits with `coord_cartesian`.  In this case, we will set it from 0 to 60. Next, we will use `scale_y_reverse()` to flip and rename the y-axis (depth). The new step here is to modify the x-axis using `scale_x_continuous()` to reflect the fact that we transformed our data with a cube-root.  The main change here is that we passed a `break =` argument to set the interval and `labels = c()` to manually control the labels. Our final step for this part is to add the `theme_plankton()` function to the end to change the overall appearance of the figure, and then save the figure to the object, `phyto_plot`

```{r}
phyto_plot <- ggplot() +
  geom_point(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 2) +
  geom_path(data = phyto_plotting, aes(x = mean_rt, y = depth), size = 1) +
  geom_errorbarh(data = phyto_plotting, aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.2) +
  facet_wrap(~taxa, nrow = 1) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Density (#/L)",
                     breaks = seq(0, 60, 20),
                     labels = c(0, expression(20^3), 
                                expression(40^3), expression(60^3))) +
  theme_plankton()
phyto_plot
```

## Adding additional elements 

If the code above worked as it was supposed to, you should see the four phytoplankton genera density plotted with depth.  We will now use the dataframes we made to add lines for the one-percent light level, the bottom of the epilimnion, and to indicate *Chaborus* presence. I'm not going to spend time going over each step.  Many of these functions should look very familiar and you should have a good enough grasp of how they work to figure out the changes for yourself! 

```{r}
phyto_plot2 <- phyto_plot +
  geom_hline(aes(yintercept = value), one_per, linetype = "dashed") +
  geom_text(data = one_per_name, aes(x = 41, y = 1.95, label = name), size = 2.5) +
  geom_hline(aes(yintercept = value), epi_bot, linetype = "dotted") +
  geom_text(data = epi_bot_name, aes(x = 41, y = 1.35, label = name), size = 2.5) +
  geom_text(data = cha_presence, aes(x = 30, y = 5, label = name), size = 4, fontface = "bold") +
  geom_text(data = cha_presence, aes(x = 30, y = 6, label = name), size = 4, fontface = "bold") +
  geom_text(data = cha_presence, aes(x = 30, y = 8, label = name), size = 4, fontface = "bold") 
phyto_plot2
```

Finally, we can save the plot with `ggsave()`.  Make sure you control the height of the figure so it doesn't look bad. 
```{r}
ggsave(phyto_plot2, file = "phyto_plot.png", height = 6)
```

# Zooplankton Plot 

The zooplankton figure is very similar to the phytoplankton figure we just made above.  The only change is that we did not transform these data with a cube-root so we do not need to modify the axis labels.  
```{r zoo plot,  eval=TRUE, echo=TRUE}
zoo_plot <- ggplot() +
  geom_point(data = zoo_plotting, aes(x = mean, y = depth), size = 2) +
  geom_path(data = zoo_plotting, aes(x = mean, y = depth), size = 1) +
  geom_errorbarh(data = zoo_plotting, aes(y = depth, xmin = min, xmax = max), height = 0.2) +
  facet_wrap(~taxa, nrow = 1) + 
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Density (#/L)") +
  theme_plankton()
zoo_plot2 <- zoo_plot +
  geom_hline(aes(yintercept = value), one_per2, linetype = "dashed") +
  geom_text(data = one_per_name2, aes(x = 41, y = 1.95, label = name), size = 2.5) +
  geom_hline(aes(yintercept = value), epi_bot2, linetype = "dotted") +
  geom_text(data = epi_bot_name2, aes(x = 41, y = 1.35, label = name), size = 2.5) +
  geom_text(data = cha_presence2, aes(x = 30, y = 5, label = name), size = 4, fontface = "bold") +
  geom_text(data = cha_presence2, aes(x = 30, y = 6, label = name), size = 4, fontface = "bold") +
  geom_text(data = cha_presence2, aes(x = 30, y = 8, label = name), size = 4, fontface = "bold") 
zoo_plot2
ggsave(zoo_plot2, file = "zoo_plot.png", height = 6)