---
title: "all_code"
author: "Cory Sauve"
date: "9/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

## Required libraries 
```{r libraries}
library(tidyverse)
library(here)
library(patchwork)
```

<<<<<<< HEAD
## Import data 
```{r import-data}
water_quality_raw <- read_csv(here("data", "water_quality.csv"))
plankton_raw <- read_csv(here("data", "plankton.csv"))
```

# WQ Data 

## Control for MDLs 
```{r mdls}
=======
# Import data 
```{r}
water_quality_raw <- read_csv(here("data/raw", "water_quality.csv"))
plankton_raw <- read_csv(here("data/raw", "plankton.csv"))
```

# Control for MDLs 
```{r}
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
# Option A 
water_quality_clean <- water_quality_raw %>% 
  mutate(
    srp_mgl = ifelse(srp_mgl <= 0.002, 0.002, srp_mgl),
    tp_mgl = ifelse(tp_mgl <= 0.002, 0.002, tp_mgl),
    no3_mgl = ifelse(no3_mgl <= 0.009, 0.009, no3_mgl),
    nh3_mgl = ifelse(nh3_mgl <= 0.015, 0.015, nh3_mgl),
    tn_mgl = ifelse(tn_mgl <= 0.104, 0.104, tn_mgl)
  )
```

<<<<<<< HEAD
## Calculating Error 
```{r error-vals}
=======
# Calculating Error 
```{r}
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
# Create custom function for error 
error_calc <- function(x){
  abs(x - lag(x))
} 

water_quality_error <- water_quality_clean %>% 
  group_by(depth) %>% 
  mutate(across(temp_c:chla_ugl, error_calc)) %>% 
  ungroup(depth) %>% 
  filter(sample_type %in% c("rep", "dup")) %>% 
  rename_at(vars(-lake_name, -depth, -sample_type), funs(paste0(., sep = "_", "error"))) %>% 
  select(-lake_name, -sample_type)
```

<<<<<<< HEAD
## Means 
```{r means}
=======
# Means 
```{r}
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
water_quality_clean <- water_quality_clean %>% 
  group_by(depth) %>% 
  summarize(across(temp_c:chla_ugl, mean, na.rm = TRUE))
```

## Organic Nitrogen 
```{r org-n}
# Create function 
get_orgn <- function(tn_mgL, no3_mgL, nh3_mgL){
<<<<<<< HEAD
    org_n <- tn_mgL - (no3_mgL + nh3_mgL)
    return(org_n)
=======
  org_n <- tn_mgL - (no3_mgL + nh3_mgL)
  return(org_n)
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
}

# Apply function to df 
water_quality_clean <- water_quality_clean %>% 
  mutate(org_n_mgl = round(get_orgn(tn_mgl, no3_mgl, nh3_mgl), 3))
```

<<<<<<< HEAD
## Percent Light Level 
```{r percent-light}
# Create function
get_percent_light <- function(light_at_depth, light_at_surface){
=======
# Percent Light Level 
```{r}
# Create function
get_percent_light <- function(light_at_depth, light_at_surface){
  
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
  percent_light <- round((light_at_depth / light_at_surface) * 100, 1)
  return(percent_light)
}

# Apply to data 
water_quality_clean <- water_quality_clean %>% 
  mutate(
    light_level_per = get_percent_light(light_dep_mmol, light_sur_mmol)
  )
```

<<<<<<< HEAD
## Sig Figs 
```{r sig-figs}
=======

# One Percent Light 
```{r}
get_one_percent <- function(depths, light){

  # Determine surface and one percent light
  surfaceLight <- light[[1]]
  onePercent <- surfaceLight * 0.01

  # Remove zeros and determine length
  light1 <- light[!light %in% 0]
  len <- length(light1)

  # Make depth vector sample length
  depths1 <- depths[1:len]

  # Calculate one percent light level
  mod <- lm(depths1 ~ log(light1))
  coef <- coef(mod)
  int <- coef[1]
  slope <- coef[2]
  onePercentLightLevel <- slope * log(onePercent) + int

  return(onePercentLightLevel)
}

water_quality_clean <- water_quality_clean %>% 
  mutate(one_percent_light = get_one_percent(depth, light_dep_mmol))
```


# Sig Figs 
```{r}
>>>>>>> b5c208dbd326da4994f376c01564bb18d60ca59a
water_quality_clean <- water_quality_clean %>% 
  mutate(
    temp_c = round(temp_c, 2), 
    do_mgl = round(do_mgl, 2),
    do_sat_per = round(do_sat_per, 1), 
    cond_umhos = round(cond_umhos, 1),
    light_sur_mmol = round(light_sur_mmol, 0),
    light_dep_mmol = round(light_dep_mmol, 0),
    ph = round(ph, 1), 
    alk_mgl = round(alk_mgl, 0),
    turb_ntu = round(turb_ntu, 1),
    srp_mgl = round(srp_mgl, 3),
    tp_mgl = round(tp_mgl, 3),
    nh3_mgl = round(nh3_mgl, 3),
    no3_mgl = round(no3_mgl, 3),
    tn_mgl = round(tn_mgl, 3),
    chla_ugl = round(chla_ugl, 2),
    org_n_mgl = round(org_n_mgl, 3),
    light_level_per = round(light_level_per, 1)
  )
```

## Combine with error values 
```{r join-error}
water_quality_clean <- water_quality_clean %>% left_join(water_quality_error, by = c("depth"))
```

## One Percent Light 
```{r one-per-light}
# Create function
get_one_percent <- function(depths, light){

  # Determine surface and one percent light
  surface_light <- light[[1]]
  one_percent <- surface_light * 0.01

  # Remove zeros and determine length
  light1 <- light[!light %in% 0]
  len <- length(light1)

  # Make depth vector sample length
  depths1 <- depths[1:len]

  # Calculate one percent light level
  mod <- lm(depths1 ~ log(light1))
  coef <- coef(mod)
  int <- coef[1]
  slope <- coef[2]
  one_percent_light_level <- slope * log(one_percent) + int

  return(one_percent_light_level)
}

# Apply to data 
one_percent_light <- round(get_one_percent(water_quality_clean$depth, water_quality_clean$light_dep_mmol), 1)
```

## Secchi 
```{r secchi}
secchi_m <- 0.75
```

## Bottom of Epi 
```{r bottom-epi}
bottom_of_epi <- 1.5
```

# Plankton Data 

## Define cube root function
```{r cube-rt}
# Define cube root function 
cube_rt <- function(x){
  x ^ (1/3)
}
```

## Transform raw plankton data 
```{r plankton-prep}
# plankton summary
plankton_summary <- plankton_raw %>% 
  group_by(depth) %>% 
  summarize(across(ends_with("nul"), list(mean = mean, min = min, max = max), .names = "{col}_{fn}")) %>% 
  pivot_longer(
    cols = dolichospermum_nul_mean:chaoborus_nul_max,
    names_to = c("taxa", "unit", "stat"),
    names_sep = "_" 
  ) %>% 
  mutate(taxa = str_to_title(taxa)) %>% 
  select(-unit) %>% 
  pivot_wider(
    names_from = stat,
    values_from = value
  )
```

## Phyt data 
```{r phyt-prep}
phyts <- plankton_summary %>% 
  filter(taxa %in% c("Dolichospermum", "Aphanizomenon", "Microcystis", "Ceratium")) %>% 
  mutate(
    mean_rt = cube_rt(mean),
    upper_rt = cube_rt(max),
    lower_rt = cube_rt(min),
    upper_bound = abs(mean_rt - upper_rt),
    lower_bound = abs(mean_rt - lower_rt)
  ) %>% 
  select(depth, taxa, mean_rt, upper_bound, lower_bound)
```

## Zoop data 
```{r zoop-prep}
zoops <- plankton_summary %>% 
  filter(!taxa %in% c("Dolichospermum", "Aphanizomenon", "Microcystis", "Ceratium")) %>% 
  rename(lower_bound = min, upper_bound = max)
```

# Figures 

## Figure 1 - Temp, DO, Turb, DO SAT 
```{r fig-1}
# Temp subplot
p_temp <- water_quality_clean %>% 
  ggplot(aes(x = temp_c, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_c_error, xmax = temp_c + temp_c_error), height = 0.1) +
  geom_segment(aes(x = 15.5, y = secchi_m, xend = 19.5, yend = secchi_m), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 0.15), label = "Secchi depth (0.75 m)", size = 3) +
  coord_cartesian(xlim = c(9, 26)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Temperature (°C)")

# DO subplot 
p_do <- water_quality_clean %>% 
  ggplot(aes(x = do_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_mgl - do_mgl_error, xmax = do_mgl + do_mgl_error), height = 0.1) +
  geom_text(aes(x = 8, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 6, y = one_percent_light, xend = 10, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 16)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "Dissolved Oxygen (mg/L)", breaks = seq(0, 16, 4))

# Turb subplot 
p_turb <- water_quality_clean %>% 
  ggplot(aes(x = turb_ntu, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = turb_ntu - turb_ntu_error, xmax = turb_ntu + turb_ntu_error), height = 0.1) +
  coord_cartesian(xlim = c(10, 40)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Turbidity (NTU)", breaks = seq(10, 40, 5))

# DO SAT subplot 
p_dosat <- water_quality_clean %>% 
  ggplot(aes(x = do_sat_per, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_sat_per - do_sat_per_error, xmax = do_sat_per + do_sat_per_error), height = 0.1) +
  coord_cartesian(xlim = c(0, 200)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "D.O. Saturation (%)", breaks = seq(0, 200, 40)) 

# Combine subplots 
plot1 <- (p_temp | p_do) / (p_turb | p_dosat) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  )

# Save plot 
ggsave(plot1, file = "figure1.png", device = "png", type = "cairo")
```

# Figure 2 - Temp, Alk, Cond, pH
```{r fig-2}
p_alk <- water_quality_clean %>% 
  ggplot(aes(x = alk_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = alk_mgl - alk_mgl_error, xmax = alk_mgl + alk_mgl_error), height = 0.1) +
  geom_text(aes(x = 153, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 140, y = one_percent_light, xend = 170, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(80, 220)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = expression("Alkalinity (mg CaCO"[3]*"/L)"))

p_cond <- water_quality_clean %>% 
  ggplot(aes(x = cond_umhos, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = cond_umhos - cond_umhos_error, xmax = cond_umhos + cond_umhos_error), height = 0.1) +
  coord_cartesian(xlim = c(280, 400)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Conductivity (μmhos/cm)", breaks = seq(280, 400, 40))

p_ph <- water_quality_clean %>% 
  ggplot(aes(x = ph, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = ph - ph_error, xmax = ph + ph_error), height = 0.1) +
  coord_cartesian(xlim = c(6, 9)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "pH", breaks = seq(6, 9, 0.5))

plot2 <- (p_temp | p_alk) / (p_cond | p_ph) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot2, file = "figure2.png", device = "png", type = "cairo")
```

# Figure 3 - Temp, NH3, Org-N, NO3
```{r fig-3}
p_nh3 <- water_quality_clean %>% 
  ggplot(aes(x = nh3_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = nh3_mgl - nh3_mgl_error, xmax = nh3_mgl + nh3_mgl_error), height = 0.2) +
  geom_text(aes(x = 8, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 6, y = one_percent_light, xend = 10, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0.0, 2.5)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = expression('NH'['3']*'-N (mg/L)'), breaks = seq(0, 2.5, .5))

p_orgn <- water_quality_clean %>% 
  ggplot(aes(x = org_n_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  coord_cartesian(xlim = c(0.0, 1.25)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Org-N (mg/L)", breaks = seq(0.0, 1.25, .25))

p_no3 <- water_quality_clean %>% 
  ggplot(aes(x = no3_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = no3_mgl - no3_mgl_error, xmax = no3_mgl + no3_mgl_error), height = 0.2) +
  coord_cartesian(xlim = c(0.005, 0.02)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = expression('NO'['3']*'-N (mg/L)'), breaks = seq(0, 0.02, 0.005))

plot3 <- (p_temp | p_nh3) / (p_orgn | p_no3) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot3, file = "figure3.png", device = "png", type = "cairo")
```

# Figure 4 - Temp, TP, SRP
```{r fig-4}
p_tp <- water_quality_clean %>% 
  ggplot(aes(x = tp_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = tp_mgl - tp_mgl_error, xmax = tp_mgl + tp_mgl_error), height = 0.1) +
  geom_text(aes(x = 8, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 6, y = one_percent_light, xend = 10, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 0.15)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "TP (mg/L)", breaks = seq(0, 0.15, 0.05))

p_srp <- water_quality_clean %>% 
  ggplot(aes(x = srp_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = srp_mgl - srp_mgl_error, xmax = srp_mgl + srp_mgl_error), height = 0.1) +
  coord_cartesian(xlim = c(0, 0.15)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "SRP (mg/L)", breaks = seq(0, 0.15, 0.05))

plot4 <- (p_temp | p_tp | p_srp) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot4, file = "figure4.png", device = "png", type = "cairo", height = 3.5, width = 8)
```

# Figure 5 - Temp, Chla 
```{r fig-5}
p_chla <- water_quality_clean %>% 
  ggplot(aes(x = chla_ugl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = chla_ugl - chla_ugl_error, xmax = chla_ugl + chla_ugl_error), height = 0.2) +
  geom_text(aes(x = 37.5, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 25, y = one_percent_light, xend = 50, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 75)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "Chlorophyll-a (μg/L)", breaks = seq(0, 75, 25))

plot5 <- (p_temp | p_chla)  + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot5, file = "figure5.png", device = "png", type = "cairo", height = 3.5, width = 5)
```

# Figure 6 - Phyt
```{r fig-6}
p_aphani <- phyts %>% 
  filter(taxa == "Aphanizomenon") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Aphanizomenon") 

p_ceratium <- phyts %>% 
  filter(taxa == "Ceratium") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Ceratium") 

p_dolicho <- phyts %>% 
  filter(taxa == "Dolichospermum") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Dolichospermum") 

p_microcystis <- phyts %>% 
  filter(taxa == "Microcystis") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_text(aes(x = 42, y = 1.9), label = "1% Light Level (2.1 m)", size = 3) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_text(aes(x = 42, y = 1.3), label = "Bottom of Epi. (1.5 m)", size = 3) +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Microcystis") 

p_phyt <- (p_aphani| p_ceratium | p_dolicho | p_microcystis) +
  plot_annotation(caption = "Density (#/L)") & 
  theme_bw() & 
  theme(plot.caption = element_text(hjust = 0.5, size = 12, vjust = 8),
        plot.title = element_text(hjust = 0.5, size = 10, face = "italic"),
        axis.title.y = element_text(size = 12),
        panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(p_phyt, file = "figure6.png", device = "png", type = "cairo", height = 6, width = 10)
```

# Figure 7 - Zoop 
```{r fig-7}
# Bosmina 
p_bosmina <- zoops %>% 
  filter(taxa == "Bosmina") %>% 
  ggplot(aes(x = mean, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = lower_bound, xmax = upper_bound), height = 0.1) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "") +
  ggtitle("Bosmina") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Calanoid 
p_calanoid <- zoops %>% 
  filter(taxa == "Calanoid") %>% 
  ggplot(aes(x = mean, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = lower_bound, xmax = upper_bound), height = 0.1) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "") +
  ggtitle("Calanoid") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Cyclopoid 
p_cyclopoid <- zoops %>% 
  filter(taxa == "Cyclopoid") %>% 
  ggplot(aes(x = mean, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = lower_bound, xmax = upper_bound), height = 0.1) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "") +
  ggtitle("Cyclopoid") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Nauplii...include b. of epi, light, chaoborus presence
p_nauplii <- zoops %>% 
  filter(taxa == "Nauplii") %>% 
  ggplot(aes(x = mean, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = lower_bound, xmax = upper_bound), height = 0.1) +
  geom_text(aes(x = 42, y = 1.9), label = "1% Light Level (2.1 m)", size = 3) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_text(aes(x = 42, y = 1.3), label = "Bottom of Epi. (1.5 m)", size = 3) +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  geom_text(aes(x = 30, y = 5), label = "X", size = 5) +
  geom_text(aes(x = 30, y = 6), label = "X", size = 5) +
  geom_text(aes(x = 30, y = 8), label = "X", size = 5) +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "") +
  ggtitle("Nauplii") +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10)
  )

p_zoop <- (p_bosmina | p_calanoid | p_cyclopoid | p_nauplii) +
  plot_annotation(caption = "Density (#/L)") &
  theme(plot.caption = element_text(hjust = 0.5, size = 12, vjust = 8))
   
ggsave(p_zoop, file = "figure7.png", device = "png", type = "cairo", height = 6, width = 10)
```

# Figure 8 - Light 
```{r fig-8}
p_light <- water_quality_clean %>% 
  ggplot(aes(x = light_level_per, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  coord_cartesian(xlim = c(0.1, 100), clip = 'off') +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_log10(name = "Light Transmittance (%)", breaks = c(0.1, 1, 10, 100), labels = c("0.01", "1", "10", "100")) +
  annotation_logticks(sides = "b") +
  geom_label(label= expression('Z'['1%']*' = 2.1m'), aes(x = 3.1, y = 2.75),
  label.padding = unit(0.55, "lines"), label.size = 0.35, color = "black", fill="white") +
  geom_segment(
    aes(x = 1.15, xend = 1.7, y = 2.1, yend = 2.7),
    arrow = arrow(ends = "first", type = "open", length = unit(0.25, "cm"))
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  )

ggsave(p_light, file = "figure8.png", device = "png", type = "cairo")
```













