---
title: "Limnology Workshop Code Outline"
author: "Cory Sauve"
date: "9/23/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting Started 

## Installing Packages 
```{r}
install.packages("rmarkdown", dependenices = TRUE)
install.packages("tidyverse", dependencies = TRUE)
install.packages("here", dependencies = TRUE)
install.packages("palmerpenguins")
```

```{r}
install.packages("tinytex")
    
# Once installed, run:
tinytex::install_tinytex()
```

# On to University Lake 

## Required Libraries 
```{r}
library(tidyverse)
library(here)
library(patchwork)
```

## Import data
```{r}
water_chem_raw <- read_csv(here("data", "water_chemistry.csv")) # Change if you named things differently 
plankton_raw <- read_csv(here("data", "plankton.csv"))
```

## Working with the water chemistry data 

### Taking a look 
```{r}
glimpse(water_chem_raw)
```

###  Control for MDL's 
```{r}
# Change MDL values with your values 
water_chem_clean <- water_chem_raw %>% 
  mutate(
    srp_mgl = ifelse(srp_mgl <= 0.002, 0.002, srp_mgl),
    tp_mgl = ifelse(tp_mgl <= 0.002, 0.002, tp_mgl),
    no3_mgl = ifelse(no3_mgl <= 0.009, 0.009, no3_mgl),
    nh3_mgl = ifelse(nh3_mgl <= 0.015, 0.015, nh3_mgl),
    tn_mgl = ifelse(tn_mgl <= 0.104, 0.104, tn_mgl)
  )
```

### Calculating Error 
```{r}
# Create function
error_calc <- function(x){
  abs(x - lag(x))
} 

# Apply function to data 
water_chem_error <- water_chem_clean %>% 
  group_by(depth) %>% 
  mutate(across(temp_c:chla_ugl, error_calc)) %>% 
  ungroup(depth) %>% 
  filter(sample_type %in% c("rep", "dup")) %>% 
  rename_at(vars(-lake_name, -depth, -sample_type), funs(paste0(., sep = "_", "error"))) %>% 
```

### Average values 
```{r}
water_chem_clean <- water_chem_clean %>% 
  group_by(depth) %>% 
  summarize(across(temp_c:chla_ugl, mean, na.rm = TRUE))
```

### Organic Nitrogen 
```{r}
# Create function
get_orgn <- function(tn_mgL, no3_mgL, nh3_mgL){
    org_n <- tn_mgL - (no3_mgL + nh3_mgL)
    return(org_n)
}

# Apply function to df 
water_chem_clean <- water_chem_clean %>% 
  mutate(orgn_mgl = round(get_orgn(tn_mgl, no3_mgl, nh3_mgl), 3))
```

### Percent Light Level 
```{r}
# Create function
get_percent_light <- function(light_at_depth, light_at_surface){
  
  percent_light <- round((light_at_depth / light_at_surface) * 100, 1)
  return(percent_light)
}

# Apply to data 
water_chem_clean <- water_chem_clean %>% 
  mutate(
    light_level_per = get_percent_light(light_dep_mmol, light_sur_mmol)
  )
```

### One Percent Light 
```{r}
# Create function
get_one_percent <- function(depths, light){

  # Determine surface and one percent light
  surface_light <- light[[1]]
  one_percent <- surface_light * 0.01

  # Remove zeros and determine length
  light1 <- light[!light %in% 0]
  len <- length(light1)

  # Make depth vector sample length
  depths1 <- depths[1:len]

  # Calculate one percent light level
  mod <- lm(depths1 ~ log(light1))
  coef <- coef(mod)
  int <- coef[1]
  slope <- coef[2]
  one_percent_light_level <- slope * log(one_percent) + int

  return(one_percent_light_level)
}

# Apply to data 
one_percent_light <- round(get_one_percent(water_chem_clean$depth, water_chem_clean$light_dep_mmol), 1)
```

### Sig Figs
```{r}
water_chem_clean <- water_chem_clean %>% 
  mutate(
    temp_c = round(temp_c, 2), 
    do_mgl = round(do_mgl, 2),
    do_sat_per = round(do_sat_per, 1), 
    cond_umhos = round(cond_umhos, 1),
    light_sur_mmol = round(light_sur_mmol, 0),
    light_dep_mmol = round(light_dep_mmol, 0),
    ph = round(ph, 1), 
    alk_mgl = round(alk_mgl, 0),
    turb_ntu = round(turb_ntu, 1),
    srp_mgl = round(srp_mgl, 3),
    tp_mgl = round(tp_mgl, 3),
    nh3_mgl = round(nh3_mgl, 3),
    no3_mgl = round(no3_mgl, 3),
    tn_mgl = round(tn_mgl, 3),
    chla_ugl = round(chla_ugl, 2),
    orgn_mgl = round(orgn_mgl, 3),
    light_level_per = round(light_level_per, 1)
  )
```

### Combining with error 
```{r}
water_chem_clean <- water_chem_clean %>% left_join(water_chem_error, by = c("depth"))
```

### Secchi 
```{r}
secchi_m <- # Update with your value 
```

### Bottom of Epilimnion
```{r}
bottom_of_epi <- # Update with your value 
```

## Working with the plankton data 

### Taking a look 
```{r}
head(plankton_raw)
```

### Defining the cube root function 
```{r}
cube_rt <- function(x){
  x ^ (1/3)
}
```

### Transforming the plankton data 
```{r}
plankton_summary <- plankton_raw %>% 
  group_by(depth) %>% 
  summarize(across(ends_with("nul"), list(mean = mean, min = min, max = max), .names = "{col}_{fn}")) %>% 
  pivot_longer(
    cols = dolichospermum_nul_mean:chaoborus_nul_max,
    names_to = c("taxa", "unit", "stat"),
    names_sep = "_" 
  ) %>% 
  mutate(taxa = str_to_title(taxa)) %>% 
  select(-unit) %>% 
  pivot_wider(
    names_from = stat,
    values_from = value
  )
```

### Phytoplankton data 
```{r}
phyts <- plankton_summary %>% 
  filter(taxa %in% c("Dolichospermum", "Aphanizomenon", "Microcystis", "Ceratium")) %>% 
  mutate(
    mean_rt = cube_rt(mean),
    upper_rt = cube_rt(max),
    lower_rt = cube_rt(min),
    upper_bound = abs(mean_rt - upper_rt),
    lower_bound = abs(mean_rt - lower_rt)
  ) %>% 
  select(depth, taxa, mean_rt, upper_bound, lower_bound)
```

### Zooplankton data
```{r}
zoops <- plankton_summary %>% 
  filter(!taxa %in% c("Dolichospermum", "Aphanizomenon", "Microcystis", "Ceratium")) %>% 
  rename(lower_bound = min, upper_bound = max)

```
 
## Making the Water Chemistry Figures 

### Figure 1 - Mapping the data 
```{r}

```

### Figure 1 - Points and Lines 
```{r}

```

### Figure 1 - Adding error bars 
```{r}

```

### Figure 1 - Adding annotations 
```{r}

```

### Figure 1 - Scale and labels 
```{r}

```

### Figure 1 - Remaining subplots 
```{r}
# DO subplot
p_do <- water_chem_clean %>% 
  ggplot(aes(x = do_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_mgl - do_mgl_error, xmax = do_mgl + do_mgl_error), height = 0.1) +
  geom_text(aes(x = 8, y = 1.5), label = "1% Light Level (2.1 m)", size = 3) +
  geom_segment(aes(x = 6, y = one_percent_light, xend = 10, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 16)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "Dissolved Oxygen (mg/L)", breaks = seq(0, 16, 4))

# Turbidity subplot 
p_turb <- water_chem_clean %>% 
  ggplot(aes(x = turb_ntu, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = turb_ntu - turb_ntu_error, xmax = turb_ntu + turb_ntu_error), height = 0.1) +
  coord_cartesian(xlim = c(10, 40)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Turbidity (NTU)", breaks = seq(10, 40, 5))

# DO SAT subplot 
p_dosat <- water_chem_clean %>% 
  ggplot(aes(x = do_sat_per, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_sat_per - do_sat_per_error, xmax = do_sat_per + do_sat_per_error), height = 0.1) +
  coord_cartesian(xlim = c(0, 200)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "D.O. Saturation (%)", breaks = seq(0, 200, 40)) 
```

### Figure 1 - Combining and adjusting 
```{r}

```

### Figure 1 - Saving
```{r}

```

### Making Figures 2-5

**LOOK AT THE END OF THIS DOCUMENT FOR TEMPLATES FOR FIGURES 2-5**

## Making the Plankton Figures 

### Figure 6 - Filtering 
```{r}
phyts %>% 
  filter()
```

### Figure 6 - Mapping the data 
```{r}
phyts %>% 
  filter() %>% 
  ggplot() 
```

### Figure 6 - Points and Lines 
```{r}
phyts %>% 
  filter() %>% 
  ggplot() + 
  geom_point() + 
  geom_path()
```

### Figure 6 - Error bars 
```{r}
phyts %>% 
  filter() %>% 
  ggplot() + 
  geom_point() + 
  geom_path()
```

### Figure 6 - More Lines 
```{r}
phyts %>% 
  filter() %>% 
  ggplot(aes()) + 
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_hline() +
  geom_hline()
```

### Figure 6 - Scale and labels 
```{r}
p_aphani <- phyts %>% 
  filter() %>% 
  ggplot() + 
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_hline(aes(), ) +
  geom_hline(aes(), ) +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous() +
  ggtitle() 
```

### Figure 6 - Remaining subplots 
```{r}
p_ceratium <- phyts %>% 
  filter(taxa == "Ceratium") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Ceratium") 

p_dolicho <- phyts %>% 
  filter(taxa == "Dolichospermum") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Dolichospermum") 

p_microcystis <- phyts %>% 
  filter(taxa == "Microcystis") %>% 
  ggplot(aes(x = mean_rt, y = depth)) + 
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = mean_rt - lower_bound, xmax = mean_rt + upper_bound), height = 0.1) +
  geom_text(aes(x = 42, y = 1.9), label = "1% Light Level (2.1 m)", size = 3) +
  geom_hline(aes(yintercept = one_percent_light), linetype = "dashed") +
  geom_text(aes(x = 42, y = 1.3), label = "Bottom of Epi. (1.5 m)", size = 3) +
  geom_hline(aes(yintercept = bottom_of_epi), linetype = "dotted") +
  coord_cartesian(xlim = c(0, 60)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(
    name = "",
    breaks = seq(0, 60, 20),
    labels = c(0, expression(20^3), expression(40^3), expression(60^3))) +
  ggtitle("Microcystis") 
```

### Figure 6 - Combining and adjusting 
```{r}
p_phyt <- (p_aphani| p_ceratium | p_dolicho | p_microcystis) +
  plot_annotation(caption = "Density (#/L)") & 
  theme_bw() & 
  theme(plot.caption = element_text(hjust = 0.5, size = 12, vjust = 8),
        plot.title = element_text(hjust = 0.5, size = 10, face = "italic"),
        axis.title.y = element_text(size = 12),
        panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))
```

### Figure 6 - Saving 
```{r}
ggsave(p_phyt, file = "figure6.png", device = "png", type = "cairo", height = 6, width = 10)
```

### Figure 7 - All at once
```{r}
p_bosmina <- zoops %>% 
  filter() %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_hline() +
  geom_hline() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous() +
  ggtitle() +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Calanoid 
p_calanoid <- zoops %>% 
  filter() %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_hline() +
  geom_hline() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous() +
  ggtitle() +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Cyclopoid 
p_cyclopoid <- zoops %>% 
  filter() %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_hline() +
  geom_hline() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous() +
  ggtitle() +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10, face = "italic")
  )

# Nauplii...include b. of epi, light, chaoborus presence
p_nauplii <- zoops %>% 
  filter() %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_text() +
  geom_hline() +
  geom_text() +
  geom_hline() +
  geom_text() +
  geom_text() +
  geom_text() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous() +
  ggtitle() +
  theme_bw() +
  theme(
    axis.title.y = element_text(size = 12),
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, size = 10)
  )

p_zoop <- (p_bosmina | p_calanoid | p_cyclopoid | p_nauplii) +
  plot_annotation(caption = "Density (#/L)") &
  theme(plot.caption = element_text(hjust = 0.5, size = 12, vjust = 8))
   
ggsave(p_zoop, file = "figure7.png", device = "png", type = "cairo", height = 6, width = 10)
```

## Making the Light Figure 

### Figure 8 - Data, Points, and Lines 
```{r}
water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() 
```

### Figure 8 - Scales and such
```{r}
water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_log10() 
```

### Figure 8 - Annotatations and arrows 
```{r}
water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_log10() +
  annotation_logticks() +
  geom_label(
    label = expression(), aes(),
    label.padding = 
  ) +
  geom_segment(
    aes(),
    arrow = arrow())
  )
```

### Figure 8 - Themes
```{r}
p_light <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_log10() +
  annotation_logticks() +
  geom_label(
    label = expression(), aes(),
    label.padding = 
  ) +
  geom_segment(
    aes(),
    arrow = arrow())
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  )
```

### Figure 8 - Saving
```{r}
ggsave(p_light, file = "figure8.png", device = "png", type = "cairo")
```

## The Rest of the Figures 

### Figure 2 - Temp, Alk, Cond, pH
```{r}
p_alk <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_text() +
  geom_segment() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

p_cond <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

p_ph <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

plot2 <- (p_temp | p_alk) / (p_cond | p_ph) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot2, file = "figure2.png", device = "png", type = "cairo")
```

### Figure 3 - Temp, NH3, Org-N, NO3
```{r}
p_nh3 <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_text() +
  geom_segment() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

p_orgn <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

p_no3 <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

plot3 <- (p_temp | p_nh3) / (p_orgn | p_no3) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot3, file = "figure3.png", device = "png", type = "cairo")
```

### Figure 4 - Temp, TP, SRP
```{r fig-4}
p_tp <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_text() +
  geom_segment() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

p_srp <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

plot4 <- (p_temp | p_tp | p_srp) + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot4, file = "figure4.png", device = "png", type = "cairo", height = 3.5, width = 8) # May need to adjust size 
```

### Figure 5 - Temp, Chla 
```{r fig-5}
p_chla <- water_chem_clean %>% 
  ggplot() +
  geom_point() + 
  geom_path() + 
  geom_errorbarh() +
  geom_text() +
  geom_segment() +
  coord_cartesian() +
  scale_y_reverse() +
  scale_x_continuous()

plot5 <- (p_temp | p_chla)  + 
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(color = "black"),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggsave(plot5, file = "figure5.png", device = "png", type = "cairo", height = 3.5, width = 5) # May need to adjust size 
```
























