---
title: "Limno Report Help"
author: "Cory Sauve"
date: "10/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

IMPORTANT REMINDERS 

- match the data template that the guide uses with your data 
- Make sure that you adjust your column names to reflect your different variables for the plankton taxa 
- Make sure that you change the depth rows to match your actual data (ie. the duplicate, replicate, split rows)
- For the figures, watch out for axis scales and label positioning.
    - For axis scales, update coord_cartesian() and breaks = seq() functions 
    - For label positioning, update the geom_text() function x= and y= 
  
  
CODE FROM OFFICE HOURS 

Loading packages you want 
```{r}
library(tidyverse)
library(here)
library(patchwork)
```

Import data 
```{r}
water_chem_raw <- read_csv(here("data", "water_chemistry.csv")) # Change if you named things differently 
plankton_raw <- read_csv(here("data", "plankton.csv"))
```

Correct for MDLs 
```{r}
water_chem_clean <- water_chem_raw %>% 
  mutate(
    srp_mgl = ifelse(srp_mgl <= 0.002, 0.002, srp_mgl),
    tp_mgl = ifelse(tp_mgl <= 0.002, 0.002, tp_mgl),
    no3_mgl = ifelse(no3_mgl <= 0.008, 0.008, no3_mgl),
    nh3_mgl = ifelse(nh3_mgl <= 0.014, 0.014, nh3_mgl),
    tn_mgl = ifelse(tn_mgl <= 0.100, 0.100, tn_mgl),
    chla_ugl = ifelse(chla_ugl <= 0.5, 0.5, chla_ugl)
  )
```

Calculate error between reps and dups
```{r}
# Create function
error_calc <- function(x){
  abs(x - lag(x))
} 

# Apply function to data 
water_chem_error <- water_chem_clean %>% 
  group_by(depth) %>% 
  mutate(across(temp_c:chla_ugl, error_calc)) %>% 
  ungroup(depth) %>% 
  filter(sample_type %in% c("rep", "dup")) %>% 
  rename_at(vars(-lake_name, -depth, -sample_type), funs(paste0(., sep = "_", "error"))) %>% 
  select(-lake_name, -sample_type)
```

Average measurements 
```{r}
water_chem_clean <- water_chem_clean %>% 
  group_by(depth) %>% 
  summarize(across(temp_c:chla_ugl, mean, na.rm = TRUE))
```

Organic Nitrogen
```{r}
# Create function
get_orgn <- function(tn_mgL, no3_mgL, nh3_mgL){
    org_n <- tn_mgL - (no3_mgL + nh3_mgL)
    return(org_n)
}

# Apply function to df 
water_chem_clean <- water_chem_clean %>% 
  mutate(orgn_mgl = round(get_orgn(tn_mgl, no3_mgl, nh3_mgl), 3))

```

Control for sig figs 
```{r}
water_chem_clean <- water_chem_clean %>% 
  mutate(
    temp_c = round(temp_c, 2), 
    do_mgl = round(do_mgl, 2),
    do_sat_per = round(do_sat_per, 1), 
    cond_umhos = round(cond_umhos, 1),
    ph = round(ph, 1), 
    alk_mgl = round(alk_mgl, 0),
    turb_ntu = round(turb_ntu, 1),
    srp_mgl = round(srp_mgl, 3),
    tp_mgl = round(tp_mgl, 3),
    nh3_mgl = round(nh3_mgl, 3),
    no3_mgl = round(no3_mgl, 3),
    tn_mgl = round(tn_mgl, 3),
    chla_ugl = round(chla_ugl, 2),
    orgn_mgl = round(orgn_mgl, 3),
    light_level_per = round(light_level_per, 1)
  )
```

Combine water chem with error values 
```{r}
water_chem_clean <- water_chem_clean %>% left_join(water_chem_error, by = c("depth"))
```

Define Secchi measurement 
```{r}
secchi_m <- 3.4
```

Defining the cube root function for phytoplankton
```{r}
cube_rt <- function(x){
  x ^ (1/3)
}
```

Transforming the plankton data 
```{r}
plankton_summary <- plankton_raw %>% 
  group_by(depth) %>% 
  summarize(across(ends_with("nul"), list(mean = mean, min = min, max = max), .names = "{col}_{fn}")) %>% 
  pivot_longer(
    cols = dolichospermum_nul_mean:chaoborus_nul_max,
    names_to = c("taxa", "unit", "stat"),
    names_sep = "_" 
  ) %>% 
  mutate(taxa = str_to_title(taxa)) %>% 
  select(-unit) %>% 
  pivot_wider(
    names_from = stat,
    values_from = value
  )
```

Getting Phytoplankton data 
```{r}
phyts <- plankton_summary %>% 
  filter(taxa %in% c("Dolichospermum", "Aphanizomenon", "Chrysosphaerella", "Dinobryon")) %>% 
  mutate(
    mean_rt = cube_rt(mean),
    upper_rt = cube_rt(max),
    lower_rt = cube_rt(min),
    upper_bound = abs(mean_rt - upper_rt),
    lower_bound = abs(mean_rt - lower_rt)
  ) %>% 
  select(depth, taxa, mean_rt, upper_bound, lower_bound)
```

Getting Zooplankton data
```{r}
zoops <- plankton_summary %>% 
  filter(!taxa %in% c("Dolichospermum", "Aphanizomenon", "Chrysosphaerella", "Dinobryon")) %>% 
  rename(lower_bound = min, upper_bound = max)
```

Defining the other variables we need 

Bottom of epilimnion: use figure below to estimate with Wetzel Fig 6-3 as guide (~PG 75))
```{r}
water_chem_clean %>% 
  ggplot(aes(x = temp_c, y = depth)) +
  geom_point(size = 1.5) +
  geom_path(size = 0.75) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_c_error, xmax = temp_c + temp_c_error), height = 0.1) +
  coord_cartesian(xlim = c(8, 30)) + 
  scale_y_reverse(name = "Depth (m)", breaks = seq(0, 8, 1)) +
  scale_x_continuous(name = "Temperature (°C)")
```

Update value below for Bottom of epi
```{r}
bottom_of_epi <- 2.8 # Change with your estimate 
```

1% Light Level
- Use the following figure to estimate (it's between 3 and 3.5 meters)
```{r}
water_chem_clean %>% 
  ggplot(aes(x = light_level_per, y = depth)) +
  geom_point() +
  geom_path() +
  coord_cartesian(xlim = c(0.1, 100), clip = "off") +
  scale_y_reverse(name = "Depth (m)", breaks = seq(0, 8, 1)) +
  scale_x_log10(
    name = "Light Transmittance (%)",
    breaks = c(0.1, 1, 10, 100),
    labels = c("0.01", "1", "10", "100")
  ) 
```

Update 1% light level value 
```{r}
one_percent_light <- 3.2 # Update with your estimate 
```

Making Figure 1 
```{r}
p_temp <- water_chem_clean %>% 
  ggplot(aes(x = temp_c, y = depth)) +
  geom_point(size = 1.5) +
  geom_path(size = 0.75) +
  geom_errorbarh(aes(y = depth, xmin = temp_c - temp_c_error, xmax = temp_c + temp_c_error), height = 0.1) +
  geom_segment(aes(x = 15, y = secchi_m, xend = 20, yend = secchi_m), size = 1, linetype = "dashed") +
  geom_text(aes(x = 17.5, y = 2.8), label = "Secchi depth (3.4 m)", size = 3) + # ADJUST Y VALUE
  coord_cartesian(xlim = c(8, 30)) + # ADJUST HERE FOR MIN/MAX VALUES
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Temperature (°C)")

p_do <- water_chem_clean %>% 
  ggplot(aes(x = do_mgl, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_mgl - do_mgl_error, xmax = do_mgl + do_mgl_error), height = 0.1) +
  geom_text(aes(x = 6, y = 2.5), label = "1% Light Level (3.2 m)", size = 3) + # ADJUST LABEL 
  geom_segment(aes(x = 4, y = one_percent_light, xend = 8, yend = one_percent_light), size = 1, linetype = "dotted") +
  coord_cartesian(xlim = c(0, 12)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "Dissolved Oxygen (mg/L)", breaks = seq(0, 16, 4))
 
p_turb <- water_chem_clean %>% 
  ggplot(aes(x = turb_ntu, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  coord_cartesian(xlim = c(0, 50)) +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_continuous(name = "Turbidity (NTU)", breaks = seq(0, 50, 10)) 
 
p_dosat <- water_chem_clean %>% 
  ggplot(aes(x = do_sat_per, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) + 
  geom_errorbarh(aes(y = depth, xmin = do_sat_per - do_sat_per_error, xmax = do_sat_per + do_sat_per_error), height = 0.1) +
  coord_cartesian(xlim = c(0, 125)) +
  scale_y_reverse(name = "") +
  scale_x_continuous(name = "D.O. Saturation (%)", breaks = seq(0, 200, 40)) 

plot1 <- (p_temp | p_do) / (p_turb | p_dosat) +
  plot_annotation(tag_levels = "A", tag_suffix = ".") &
  theme_bw() &
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  )

plot1
ggsave(plot1, file = "figure1.png", device = "png", type = "cairo", width = 7, height = 7)
```

Making Figure 8 - Light 
```{r}
p_light <- water_chem_clean %>% 
  ggplot(aes(x = light_level_per, y = depth)) +
  geom_point(size = 1.5) + 
  geom_path(size = 0.75) +
  coord_cartesian(xlim = c(0.1, 100), clip = "off") +
  scale_y_reverse(name = "Depth (m)") +
  scale_x_log10(
    name = "Light Transmittance (%)",
    breaks = c(0.1, 1, 10, 100),
    labels = c("0.01", "1", "10", "100")
  ) +
  annotation_logticks(sides = "b") +
  geom_label(
    label = expression('Z'['1%']*' = 3.2m'), aes(x = 3.0, y = 4), # Update with your value 
    label.padding = unit(0.55, "lines"), label.size = 0.35, color = "black", fill = "white"
  ) +
  geom_segment(
    aes(x = 1.15, xend = 1.7, y = 3.1, yend = 3.7),
    arrow = arrow(ends = "first", type = "open", length = unit(0.25, "cm"))
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(), 
    panel.border = element_rect(color = "black"),
    axis.text = element_text(color = "black"),
    axis.ticks = element_line(color = "black")
  )

p_light
ggsave(p_light, file = "figure8.png", device = "png", type = "cairo")

```


